---
title: "BioCor"
author: "[Lluís Revilla](mailto:lrevilla@clinic.cat)"
date: "`r doc_date()`"
package: "`r pkg_ver('BioCor')`"
output: 
  BiocStyle::html_document:
    toc: yes
    keep_md: true
    fig_caption: yes
  BiocStyle::pdf_document:
    toc: yes
    fig_caption: yes
    keep_md: true
vignette: >
  %\VignetteIndexEntry{BioCor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

# BioCor Vignette

This Vignette describes what is the purpose of this package and how to use it.

## Citation

## Quick start

To speed calculation we need a parallel structure, (or very few genes):

```{r simple_start, collapse = TRUE}
library("BioCor")
library("foreach")
library("doParallel")
registerDoParallel(2)
# GO still gives problems sometimes
genes.id <- as.character(c(52, 11342, 80895, 57654, 58493, 1164, 1163, 4150, 
                           2130, 159))
similarities <- bioCor(genes.id, kegg = TRUE, react = TRUE, go = FALSE)
stopImplicitCluster()
similarities
```
Now in similarity we have a list of matrices of `r ncol(similarities[[1]])` rows and columns. 

## Deeper understanding
### Dealing with unconclusive Ids
How does it deal when one ID is mapped to several ones?

## WGCNA and BioCor 

BioCor, is designed to be used with other clustering packages using similarity measures like WGCNA, does. Here we provide an example on how to use BioCor to improve the modules using WGCNA:

## Session Info
```{r session}
sessionInfo()
```
